Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-18 paper.qmd
Added model interpretations for MNIST digit 1 data set

## Decide by looking at MSE plot
num_bins_x_tsne_gau <- 8
num_bins_y_tsne_gau <- 10
hex_size_tsne_gau <- 0.1

@@

# hex_size_vec <- seq(0.02, 2, by = 0.01)
# 
# for (i in 1:length(hex_size_vec)) {
#   
#   find_non_empty_bins(data = umap_gau_scaled, x = "UMAP1", y = "UMAP2",
#                     non_empty_bins = 34, x_start = NA, y_start = NA,
#                     buffer_x = NA, buffer_y = NA, hex_size = hex_size_vec[i],
#                     col_start = "UMAP")
#   
#   
# }

@@

@@

class_avg <- umap_pbmc_scaled |>
  group_by(cell_label) |>
  summarise(
    UMAP1 = median(UMAP1),
    UMAP2 = median(UMAP2)
  ) |>
  mutate(cell_label = as.numeric(cell_label)) |>
  arrange(cell_label)

class_avg$cell_label <- as.factor(class_avg$cell_label)

levels(class_avg$cell_label) <- c("Memory \nCD4 T", "Naive CD4 T", "CD14+ Mono",  "B", "CD8 T", "FCGR3A+ \n Mono", "NK", "M-MDSC\n-like", "CD27-CD+ \n Memory T", "DC")

plot_list2_pbmc <- umap_pbmc_scaled |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = cell_label))+
    geom_point(alpha=0.5) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#b15928", "#1f78b4", "#cab2d6", "#ccebc5", "#fb9a99", "#e31a1c", "#6a3d9a", "#ff7f00", "#ffed6f", "#fdbf6f", "#ffff99", "#a6cee3", "#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#b2df8a", "#bc80bd", "#33a02c", "#ccebc5", "#ffed6f", "#000000", "#bdbdbd")) +
  geom_text(aes(x=UMAP1,y = UMAP2,label = cell_label), data = class_avg,inherit.aes = F, color = "black",fontface = "bold",size = 2) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) 
```

<!--obtain 2D and high-D distances-->

<!---
```{r}
#| warning: false
#| echo: false
#| message: false
#| eval: false

## Obtian 2D distances
proxy::dist(umap_pbmc[, -3]) |> 
as.data.frame()

## Obtain high-D distances
proxy::dist(training_data_pbmc[, -16]) |> 
as.data.frame()
```
--->

@@

## To draw with AIC
aic_plot_pbmc <- ggplot(mse_df_pbmc, aes(x = num_bins, y = aic
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 80, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("AIC") +
  xlab("total number of bins")
## Effective number of bins along x-axis

mse_plot_pbmc <- ggplot(mse_df_pbmc, aes(x = num_bins,
                                       y = mse
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 80, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("MSE") +
  xlab("total number of bins")

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training PBMC3k dataset. What is the effective number of bins in each NLDR technique to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 80, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-diagnosticpltPBMC
##| out-width: 100%
#| fig-pos: H

@@

## Decide by looking at MSE plot
num_bins_x_pbmc <- 8
num_bins_y_pbmc <- 10
hex_size_pbmc <- 0.1

@@

trimesh_removed_pbmc_umap <- trimesh_removed_pbmc_umap +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

plot_list2_pbmc_error <- error_df |>
  mutate(type = factor(type , levels = c("no error", "error 1-15", "error 15-20", "error 20-25", "error 25-30", "error 30-35", "error 35-40", "error greter than 40"))) |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = type))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#fee0d2", "#fcbba1",
                                         "#fc9272", "#fb6a4a", "#ef3b2c",
                                         "#cb181d", "#a50f15", "#99000d")) 
```

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 30, min_dist = 0.3) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<https://youtu.be/CNXjWZMyh_M>)"
#| label: fig-nldervisPBMCUMAP
#| out-width: 100%
#| fig-pos: h

plot_list2_pbmc + trimesh_removed_pbmc_umap + plot_list2_pbmc_error +
plot_layout(guides='collect', ncol=3) &
  theme(legend.position='bottom')
```

@@

class_avg <- umap_pbmc_scaled |>
  group_by(cell_label) |>
  summarise(
    UMAP1 = median(UMAP1),
    UMAP2 = median(UMAP2)
  ) |>
  mutate(cell_label = as.numeric(cell_label)) |>
  arrange(cell_label)

class_avg$cell_label <- as.factor(class_avg$cell_label)

levels(class_avg$cell_label) <- c("Memory \nCD4 T", "Naive CD4 T", "CD14+ Mono",  "B", "CD8 T", "FCGR3A+ \n Mono", "NK", "M-MDSC\n-like", "CD27-CD+ \n Memory T", "DC")

plot_list2_pbmc <- umap_pbmc_scaled |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = cell_label))+
    geom_point(alpha=0.5) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#b15928", "#1f78b4", "#cab2d6", "#ccebc5", "#fb9a99", "#e31a1c", "#6a3d9a", "#ff7f00", "#ffed6f", "#fdbf6f", "#ffff99", "#a6cee3", "#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#b2df8a", "#bc80bd", "#33a02c", "#ccebc5", "#ffed6f", "#000000", "#bdbdbd")) +
  geom_text(aes(x=UMAP1,y = UMAP2,label = cell_label), data = class_avg,inherit.aes = F, color = "black",fontface = "bold",size = 2) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) 
```

@@

## Decide by looking at MSE plot
num_bins_x_pbmc <- 8
num_bins_y_pbmc <- 11
hex_size_pbmc <- 0.1

@@

trimesh_removed_pbmc_umap <- trimesh_removed_pbmc_umap +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

plot_list2_pbmc_error <- error_df |>
  mutate(type = factor(type , levels = c("no error", "error 1-15", "error 15-20", "error 20-25", "error 25-30", "error 30-35", "error 35-40", "error greter than 40"))) |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = type))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#fee0d2", "#fcbba1",
                                         "#fc9272", "#fb6a4a", "#ef3b2c",
                                         "#cb181d", "#a50f15", "#99000d")) 
```

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 5, min_dist = 0.99, metric = cosine) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<>)"
#| label: fig-nldervisPBMCUMAP3
#| out-width: 100%
#| fig-pos: h

plot_list2_pbmc + trimesh_removed_pbmc_umap + plot_list2_pbmc_error +
plot_layout(guides='collect', ncol=3) &
  theme(legend.position='bottom')

@@

```{r}
#| echo: false
#| fig-cap: "2D layout from UMAP (n_neighbors = 5, min_dist = 0.99, metric = cosine) applied for PBMC3k dataset (a) colored manually in 2D, (b) colored in high-D"
#| label: fig-nldervisPBMCUMAP3Res
#| out-width: 100%

@@

@@

- Learned from the model: There is a non-linear shape in 2D, but if look at the model in high-D, the non-linear continuity is there but there is a twist in the structure which didn't capture by PaCMAP.

>>>>

## Decide by looking at MSE plot
num_bins_x_tsne_gau <- 21
num_bins_y_tsne_gau <- 28
hex_size_tsne_gau <- 0.03
## non-empty:198

@@

## Decide by looking at MSE plot
num_bins_x_tsne_gau <- 60
num_bins_y_tsne_gau <- 79
hex_size_tsne_gau <- 0.01
### non-empty 229

hb_obj_tsne_gau <- hex_binning(data = umap_gau_scaled, x = "UMAP1", 
                      y = "UMAP2", num_bins_x = num_bins_x_tsne_gau, 
                      num_bins_y = num_bins_y_tsne_gau, x_start = NA, y_start = NA, 
                      buffer_x = NA, buffer_y = NA, hex_size = hex_size_tsne_gau, col_start = "UMAP")

model_object_tsne_gau <- fit_highd_model( training_data = training_data_gau, 
                                 nldr_df_with_id = umap_gau_scaled, 
                                 x = "UMAP1", y = "UMAP2", 
                                 num_bins_x = num_bins_x_tsne_gau, 
                                 num_bins_y = num_bins_y_tsne_gau, 
                                 x_start = NA, y_start = NA, 
                                 buffer_x = NA, buffer_y = NA, 
                                 hex_size = hex_size_tsne_gau,
                                 is_rm_lwd_hex = FALSE, 
                                 benchmark_to_rm_lwd_hex = NA, 
                                 col_start_2d = "UMAP", 
                                 col_start_highd = "x")

# show_langevitour(df = df_all_tsne_gau, df_b = df_bin_tsne_gau, df_b_with_center_data = df_bin_centroids_tsne_gau, col_start = "x", distance_df = distance_tsne_gau, distance_col = "distance", benchmark_value = benchmark_tsne_gau)

  # find_non_empty_bins(data = umap_gau_scaled, x = "UMAP1", y = "UMAP2",
  #                   non_empty_bins = 198, x_start = NA, y_start = NA,
  #                   buffer_x = NA, buffer_y = NA, hex_size = 0.03,
  #                   col_start = "UMAP")

# hex_size_vec <- seq(0.02, 2, by = 0.01)
# 
# for (i in 1:length(hex_size_vec)) {
# 
#   find_non_empty_bins(data = umap_gau_scaled, x = "UMAP1", y = "UMAP2",
#                     non_empty_bins = 198, x_start = NA, y_start = NA,
#                     buffer_x = NA, buffer_y = NA, hex_size = hex_size_vec[i],
#                     col_start = "UMAP")
# 
# 
# }

```{r}
#| warning: false
#| echo: false
#| message: false
#| eval: false

phate_gau_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = PHATE_data_gau, x = "PAHTE1", y = "PAHTE2"))) |>
  dplyr::rename(c("PAHTE1" = "scaled_PAHTE1", 
                  "PAHTE2" = "scaled_PAHTE2")) |>
  dplyr::mutate(ID = 1:NROW(PAHTE_data_gau))

## Decide by looking at MSE plot
num_bins_x_tsne_gau <- 46
num_bins_y_tsne_gau <- 61
hex_size_tsne_gau <- 0.013
### non-empty 179

hb_obj_tsne_gau <- hex_binning(data = phate_gau_scaled, x = "PHATE1", 
                      y = "PHATE2", num_bins_x = num_bins_x_tsne_gau, 
                      num_bins_y = num_bins_y_tsne_gau, x_start = NA, y_start = NA, 
                      buffer_x = NA, buffer_y = NA, hex_size = hex_size_tsne_gau, col_start = "PHATE")

model_object_tsne_gau <- fit_highd_model( training_data = training_data_gau, 
                                 nldr_df_with_id = phate_gau_scaled, 
                                 x = "PHATE1", y = "PHATE2", 
                                 num_bins_x = num_bins_x_tsne_gau, 
                                 num_bins_y = num_bins_y_tsne_gau, 
                                 x_start = NA, y_start = NA, 
                                 buffer_x = NA, buffer_y = NA, 
                                 hex_size = hex_size_tsne_gau,
                                 is_rm_lwd_hex = FALSE, 
                                 benchmark_to_rm_lwd_hex = NA, 
                                 col_start_2d = "PHATE", 
                                 col_start_highd = "x")

# show_langevitour(df = df_all_tsne_gau, df_b = df_bin_tsne_gau, df_b_with_center_data = df_bin_centroids_tsne_gau, col_start = "x", distance_df = distance_tsne_gau, distance_col = "distance", benchmark_value = benchmark_tsne_gau)

trimap_gau_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = TriMAP_data_gau, 
                                    x = "TriMAP1", y = "TriMAP2"))) |>
  dplyr::rename(c("TriMAP1" = "scaled_TriMAP1", 
                  "TriMAP2" = "scaled_TriMAP2")) |>
  dplyr::mutate(ID = 1:NROW(TriMAP_data_gau))

## Decide by looking at MSE plot
num_bins_x_tsne_gau <- 46
num_bins_y_tsne_gau <- 61
hex_size_tsne_gau <- 0.013
### non-empty 179

hb_obj_tsne_gau <- hex_binning(data = trimap_gau_scaled, x = "TriMAP1", 
                      y = "TriMAP2", num_bins_x = num_bins_x_tsne_gau, 
                      num_bins_y = num_bins_y_tsne_gau, x_start = NA, y_start = NA, 
                      buffer_x = NA, buffer_y = NA, hex_size = hex_size_tsne_gau, col_start = "TriMAP")

model_object_tsne_gau <- fit_highd_model( training_data = training_data_gau, 
                                 nldr_df_with_id = trimap_gau_scaled, 
                                 x = "TriMAP1", y = "TriMAP2", 
                                 num_bins_x = num_bins_x_tsne_gau, 
                                 num_bins_y = num_bins_y_tsne_gau, 
                                 x_start = NA, y_start = NA, 
                                 buffer_x = NA, buffer_y = NA, 
                                 hex_size = hex_size_tsne_gau,
                                 is_rm_lwd_hex = FALSE, 
                                 benchmark_to_rm_lwd_hex = NA, 
                                 col_start_2d = "TriMAP", 
                                 col_start_highd = "x")

# show_langevitour(df = df_all_tsne_gau, df_b = df_bin_tsne_gau, df_b_with_center_data = df_bin_centroids_tsne_gau, col_start = "x", distance_df = distance_tsne_gau, distance_col = "distance", benchmark_value = benchmark_tsne_gau)

pacmap_gau_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = PaCMAP_data_gau, 
                                    x = "PaCMAP1", y = "PaCMAP2"))) |>
  dplyr::rename(c("PaCMAP1" = "scaled_PaCMAP1", 
                  "PaCMAP2" = "scaled_PaCMAP2")) |>
  dplyr::mutate(ID = 1:NROW(PaCMAP_data_gau))

## Decide by looking at MSE plot
num_bins_x_tsne_gau <- 31
num_bins_y_tsne_gau <- 40
hex_size_tsne_gau <- 0.02
### non-empty 179

hb_obj_tsne_gau <- hex_binning(data = pacmap_gau_scaled, x = "PaCMAP1", 
                      y = "PaCMAP2", num_bins_x = num_bins_x_tsne_gau, 
                      num_bins_y = num_bins_y_tsne_gau, x_start = NA, y_start = NA, 
                      buffer_x = NA, buffer_y = NA, hex_size = hex_size_tsne_gau, col_start = "PaCMAP")

model_object_tsne_gau <- fit_highd_model( training_data = training_data_gau, 
                                 nldr_df_with_id = pacmap_gau_scaled, 
                                 x = "PaCMAP1", y = "PaCMAP2", 
                                 num_bins_x = num_bins_x_tsne_gau, 
                                 num_bins_y = num_bins_y_tsne_gau, 
                                 x_start = NA, y_start = NA, 
                                 buffer_x = NA, buffer_y = NA, 
                                 hex_size = hex_size_tsne_gau,
                                 is_rm_lwd_hex = FALSE, 
                                 benchmark_to_rm_lwd_hex = NA, 
                                 col_start_2d = "PaCMAP", 
                                 col_start_highd = "x")

# show_langevitour(df = df_all_tsne_gau, df_b = df_bin_tsne_gau, df_b_with_center_data = df_bin_centroids_tsne_gau, col_start = "x", distance_df = distance_tsne_gau, distance_col = "distance", benchmark_value = benchmark_tsne_gau)

@@

### Story

- UMAP representation of PBMC3k data with 15 PCs and parameter choice of n:neighbor:30, min_dist:0.1, metric:cosine (author's suggestion) shows three well-separated clusters. The question we need to answer is: Does this UMAP representation preserve the actual data structure, or is something missing or misidentified or misleading?

@@

plot_list2_pbmc <- umap_pbmc_scaled |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.5) +
  coord_equal() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
        legend.key.height = unit(0.25, 'cm'),
        legend.key.width = unit(0.25, 'cm')) #+
  # scale_color_manual(values=c("#b15928", "#1f78b4", "#cab2d6", "#ccebc5", "#fb9a99", "#e31a1c", "#6a3d9a", "#ff7f00", "#ffed6f", "#fdbf6f", "#ffff99", "#a6cee3", "#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#b2df8a", "#bc80bd", "#33a02c", "#ccebc5", "#ffed6f", "#000000", "#bdbdbd")) +
  # geom_text(aes(x=UMAP1,y = UMAP2,label = cell_label), data = class_avg,inherit.aes = F, color = "black",fontface = "bold",size = 2) +
  #annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) 
```

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 30, min_dist = 0.3) applied for the PBMC3k dataset. Is this a best representation of the original data?"
#| label: fig-nldervisPBMCUMPview
#| fig-pos: H

plot_list2_pbmc 
```

- To investigate this, we are going to visualise the model in the data space, which means that the model generated from UMAP is visualised in high-D with original data.

- To fit the model, first we need to decide how many number bins should allocate along the x and y axes. To make an initial choice, we use MSE plot.

@@

## To draw with AIC
aic_plot_pbmc <- ggplot(mse_df_pbmc, aes(x = num_bins, y = aic
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 588, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("AIC") +
  xlab("total number of bins")
## Effective number of bins along x-axis

mse_plot_pbmc <- ggplot(mse_df_pbmc, aes(x = num_bins,
                                       y = mse
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 588, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("MSE") +
  xlab("total number of bins")

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training PBMC3k dataset. What is the effective number of bins in each NLDR technique to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 588, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-diagnosticpltPBMC
##| out-width: 100%
#| fig-pos: H

@@

- Then, the model is constructed by following all the steps in 2D and lifts the model into high-D.

## Decide by looking at MSE plot
num_bins_x_pbmc <- 21
num_bins_y_pbmc <- 28
hex_size_pbmc <- 0.03

@@

trimesh_removed_pbmc_umap <- trimesh_removed_pbmc_umap +
  geom_point(data = umap_pbmc_scaled, aes(x = UMAP1, y = UMAP2), alpha = 0.3) +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) #+
  #annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

# df_all_mnist <- dplyr::bind_cols(training_data_pbmc |> dplyr::select(-ID), umap_data_with_hb_id_pbmc)
# 
# show_langevitour(df_all_pbmc, df_bin_pbmc, df_bin_centroids_pbmc, benchmark_pbmc, distance_pbmc, "distance", col_start = "PC_")

# df_bin_train <- df_bin_pbmc
# names(df_bin_train)[-1] <- paste0("avg_", names(df_bin_train)[-1])
# 
# df_all_pbmc <- dplyr::bind_cols(training_data_pbmc |> dplyr::select(-ID), umap_data_with_hb_id_pbmc)
# 
# error_df <- df_all_pbmc |>
#   dplyr::left_join(df_bin_train, by = c("hb_id" = "hb_id")) ## Map high-D averaged/weighted mean coordinates
# 
# for (i in 1:(NCOL(df_bin_train) - 1)) {
#   
#   error_df[ , paste0("abs_residual_", "PC_", i)] <- abs(error_df[ , paste0("PC_", i)] - error_df[ , paste0("avg_", "PC_", i)])
#   
# }
# 
# error_df <- error_df |>
#   dplyr::mutate(total = rowSums(dplyr::pick(tidyselect::starts_with(paste0("abs_residual_", "PC_")))))
# 
# 
# error_df <- error_df |>
#   mutate(type = if_else(total == 0, "no error",
#                         if_else(total <= 15, "error 1-15",
#                                 if_else(total <= 20, "error 15-20",
#                                         if_else(total <= 25, "error 20-25",
#                                                 if_else(total <= 30, "error 25-30",
#                                                         if_else(total <= 35, "error 30-35",
#                                                                 if_else(total <= 40, "error 35-40", "error greter than 40"))))))))
# 
# plot_list2_pbmc_error <- error_df |>
#   mutate(type = factor(type , levels = c("no error", "error 1-15", "error 15-20", "error 20-25", "error 25-30", "error 30-35", "error 35-40", "error greter than 40"))) |>
#     ggplot(aes(x = UMAP1,
#                y = UMAP2, color = type))+
#     geom_point(alpha=0.5, size = 0.1) +
#     coord_equal() +
#     theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
#   theme_linedraw() +
#     theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
#               axis.title.x = element_blank(), axis.title.y = element_blank(),
#               axis.text.x = element_blank(), axis.ticks.x = element_blank(),
#               axis.text.y = element_blank(), axis.ticks.y = element_blank(),
#               panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
#         legend.title = element_text(size=5), #change legend title font size
#         legend.text = element_text(size=4),
#          legend.key.height = unit(0.25, 'cm'),
#          legend.key.width = unit(0.25, 'cm')) +
#   scale_color_manual(values=c("#fee0d2", "#fcbba1",
#                                          "#fc9272", "#fb6a4a", "#ef3b2c",
#                                          "#cb181d", "#a50f15", "#99000d")) 
```

```{r}
#| echo: false
#| fig-cap: " Model in the 2D space with UMAP (<https://youtu.be/CNXjWZMyh_M>)"
#| label: fig-nldervisPBMCUMAP
#| fig-pos: H

trimesh_removed_pbmc_umap 
```

<!--add screen shots of model in high-D-->

- You will see three well-separated clusters in high-D as well which gives evidence that UMAP preserves the global structure. Furthermore, there is non-linearity within the clusters which was not identified by the UMAP, which proves that local structure is not preserved.

- Can we find a better parameter choice for UMAP which preserve the local structure as well?

- To do that, we are going to fit models for different parameter choice and select the one which have the lowest error.

- Selected the one with n:neighbor:5, min_dist:0.99, metric:cosine.

- Now, we are going to fit the model for the new parameter choice.

@@

plot_list2_pbmc <- umap_pbmc_scaled |>
    ggplot(aes(x = UMAP1,
               y = UMAP2))+
    geom_point(alpha=0.5) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) 
```

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 5, min_dist = 0.99, metric = cosine) applied for the PBMC3k dataset. Is this a best representation of the original data?"
#| label: fig-nldervisPBMCUMAPview2
#| fig-pos: H

plot_list2_pbmc

@@

## Decide by looking at MSE plot
num_bins_x_pbmc <- 21
num_bins_y_pbmc <- 28
hex_size_pbmc <- 0.03

@@

trimesh_removed_pbmc_umap <- trimesh_removed_pbmc_umap +
  geom_point(data = umap_pbmc_scaled, aes(x = UMAP1, y = UMAP2), alpha = 0.3) +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

plot_list2_pbmc_error <- error_df |>
  mutate(type = factor(type , levels = c("no error", "error 1-15", "error 15-20", "error 20-25", "error 25-30", "error 30-35", "error 35-40", "error greter than 40"))) |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = type))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#fee0d2", "#fcbba1",
                                         "#fc9272", "#fb6a4a", "#ef3b2c",
                                         "#cb181d", "#a50f15", "#99000d")) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

# show_langevitour(df_all_pbmc, df_bin_pbmc, df_bin_centroids_pbmc, benchmark_pbmc, distance_pbmc, "distance", col_start = "PC_")
```

```{r}
#| echo: false
#| fig-cap: "Model in the 2D space with UMAP (<>)"
#| label: fig-nldervisPBMCUMAP3
#| out-width: 100%
#| fig-pos: H

trimesh_removed_pbmc_umap + plot_list2_pbmc_error +
plot_layout(guides='collect', ncol=2) &
  theme(legend.position='bottom')

@@

```{r}
#| echo: false
#| fig-cap: "2D layout from UMAP (n_neighbors = 5, min_dist = 0.99, metric = cosine) applied for PBMC3k dataset (a) colored manually in 2D, (b) colored in high-D"
#| label: fig-nldervisPBMCUMAP3Res
#| out-width: 100%
#| fig-pos: H

@@

- Why I choose this dataset? Because I need show that whatever NLDR technique that we use, our model can capture the non-linear structure within clusters as well.

@@

- Learned from the model: There is a non-linear shape in 2D, but if look at the model in high-D, the non-linear continuity is there but there is a twist in the structure which didn't capture by PaCMAP in 2D layout.

#### Story
